Does the COMPAS needle always point towards equity? Finding Fairness in the COMPAS Risk Assessment Algorithm

### 2022-03-08 Paper_Draft1/Draft1.tex

%=================================================================
% Full title of the paper (Capitalized)
\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
Assessment Algorithm?}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private for-profit company
Northpointe, also known by its parent company Equivant
\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
It generates a score that classifies defendants' risk of recidivism as
either low, medium, or high \citep{angwin2016machine}. Jurisdictions
across the United States use the COMPAS risk assessment instrument. In
2016, ProPublica published a piece that analyzed the methods and
algorithms used by Northpointe to uncover racial biases in defendants'
scores \citep{angwin2016machine}. Their analysis looks at the
distribution of decile COMPAS scores among Black and white defendants.
ProPublica concludes, after using a statistical parity metric on false
positives or false negatives, that the algorithm is racially biased
\citep{larson2016we}. Northpointe denies the allegations of racial bias,
conducting their own analyses based on different statistical parity
metrics \citep{equivant_response_2018}. Since then, the two parties have
had several exchanges, maintaining their original arguments. Currently,
the two parties' disagreement focuses on pretrial Risk Assessment
Instruments (RAIs). ProPublica maintains that there are biases in the
outcome values, protected attributes, and covariates during
Northpointe's data processing phase. ProPublica accounts for these
biases in their analyses, and Northepointe's response, highlights how
ProPublica did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

@@

\hypertarget{bibliography}{%
\section{Bibliography}\label{bibliography}}

\citet{angwin2016machine} \citet{bao2021s} \citet{barocas2017fairness}
\citet{baumer2017texts} \citet{equivant_response_2018}
\citet{gebru2021datasheets} \citet{hardin2015data}
\citet{james2013introduction} \citet{knight2020automated}
\citet{kypraiou_what_2021} \citet{larson2016technical}
\citet{larson2016we} \citet{noauthor_ai_nodate}
\citet{noauthor_aif360datasetscompasdataset_2018}
\citet{noauthor_trusted-ai_360}
\citet{noauthor_propublicacompas-analysis_2022}
\citet{noauthor_womenthetable_nodate} \citet{vartan_1st_2022}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Does the COMPAS needle always point towards equity? Finding
Fairness in the COMPAS Risk Assessment Algorithm}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private for-profit company
Northpointe, also known by its parent company Equivant
\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
It generates a score that classifies defendants' risk of recidivism as
either low, medium, or high \citep{angwin2016machine}. Jurisdictions
across the United States use the COMPAS risk assessment instrument. In
2016, ProPublica published a piece that analyzed the methods and
algorithms used by Northpointe to uncover racial biases in defendants'
scores \citep{angwin2016machine}. Their analysis looks at the
distribution of decile COMPAS scores among Black and white defendants.
ProPublica concludes, after using a statistical parity metric on false
positives or false negatives, that the algorithm is racially biased
\citep{larson2016we}. Northpointe denies the allegations of racial bias,
conducting their own analyses based on different statistical parity
metrics \citep{equivant_response_2018}. Since then, ProPublica has
responded, maintaining their original argument and Northpointe has yet
to respond (though in 2017, they adopted the name Equivant). Currently,
the two parties' disagreement focuses on pretrial Risk Assessment
Instruments (RAIs). ProPublica maintains that there are biases in the
outcome values, protected attributes, and covariates during
Northpointe's data processing phase. ProPublica accounts for these
biases in their analyses, and Northepointe's response, highlights how
ProPublica did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

@@
